$R^*$: A robust MCMC convergence diagnostic with uncertainty via gradient-boosted machines

### 2020-03-16 write_up/ms.tex
first draft of intro

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution. Collectively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if this predictions are accurate, then the chains have not mixed and convergence has not occurred. Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$. In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain. By taking advantage of the uncertainty in chain predictions from the fitted machine learning model, our method also produces a measure of uncertainty in $R^*$. The method is straightforward to implement, robust to machine learning hyperparameter choice, and could be a complimentary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through its many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpinned the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags}, to the more recent, and also popular, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no} that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement. MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care. A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised depictions of the posterior. Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling. Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use. We repeat here a quote from Chris Holmes to this effect \cite{holmes},

@@

The predominant diagnostic method for determining whether practical convergence has occurred relies on the idea that the posterior distribution is the unique stationary distribution for an MCMC sampler. Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred. Unfortunately, anyone who uses MCMC knows that the method is full of false dawns: the Markov chains can easily become stuck in areas of parameter space and observation over short intervals can mean that the sampling distribution \textit{appears} converged. Like furious bees trapped in a room of a house [cite video], MCMC samplers may fail to move because of narrow gaps joining neighbouring areas. With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.

\section{Method}

\section{Results}
To illustrate the utility of our approach, we use examples taken from \cite{vehtari2019rank}.

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
	\caption{\textbf{Autoregressive example.} A shows how the gradient-boosted model's classifications vary according to the sample's value for an example model fit; B shows the ratio of predictive accuracy to the null model accuracy across 1000 replicates; C shows the uncertainty in accuracy ratio for two series described in the text. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
	\label{fig:ar1}
\end{figure}

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
	\label{fig:cauchy}
\end{figure}

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains; B shows the same but using the 4 original chains. In both cases, the plots show 1000 samples. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
	\label{fig:eight_schools}
\end{figure}

>>>>

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution. Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred. Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$. In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain. Since our machine learning method, gradient-boosted regression trees, provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to machine learning hyperparameter choice, and could be a complimentary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to the more recent algorithms: Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement. MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care. A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior. Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling. Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use. We repeat here a quote from Chris Holmes to this effect \cite{holmes},

@@

The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler. Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred. Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space and observation over short intervals mean the sampling distribution \textit{appears} converged. Like furious bees trapped in a room of a house \cite{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas. With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.

To combat this curse of hindsight, running multiple, independent, chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard recommended practice \cite{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \cite{gelman2013bayesian}. If these variances are similar, $\hat{R}\approx 1$ and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves \cite{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \cite{vehtari2019rank} and adopted \cite{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if the chains are mixed, it should not be possible to discern from a sample the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that caused each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across all quantiles and parameters -- any of which signal that convergence hasn't occurred. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$ whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \cite{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \cite{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated and is insensitive to the GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since the outputs from GBMs are class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful status of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries and we argue provides a complementary convergence metric alongside more established measures.

The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank}. Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.

%$R^*$ calculation is enabled in Pints inference software \cite{Clerx2019Pints}.

\section{Method}\label{sec:method}

\section{Results}\label{sec:results}
To illustrate the utility of our approach, we use examples taken from \cite{vehtari2019rank}.

@@

%\begin{figure}[h]
%	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
%	\caption{\textbf{Autoregressive example.} A shows how the gradient-boosted model's classifications vary according to the sample's value for an example model fit; B shows the ratio of predictive accuracy to the null model accuracy across 1000 replicates; C shows the uncertainty in accuracy ratio for two series described in the text. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
%	\label{fig:ar1}
%\end{figure}

@@

%\begin{figure}[h]
%	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
%	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
%	\label{fig:cauchy}
%\end{figure}

@@

%\begin{figure}[h]
%	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
%	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains; B shows the same but using the 4 original chains. In both cases, the plots show 1000 samples. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
%	\label{fig:eight_schools}
%\end{figure}
